---
title: "What SVMC finds: mutually exclusive inversions predict Salmonella serovars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What SVMC finds: mutually exclusive inversions predict Salmonella serovars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5, dpi = 150)
```

## The question

Bacterial chromosomes undergo large-scale structural rearrangements -- most
prominently, inversions spanning hundreds of kilobases to megabases. A natural
hypothesis is that such large inversions might mark distinct sublineages of a
species. **SVMC** (Structural Variant Mapping and Characterization) was built to
test this directly: it calls structural variants from complete genome
assemblies, groups recurrent ones into motifs, and asks which -- if any -- are
lineage-defining.

The answer, worked through below on *Salmonella enterica*, is more specific than
the original hypothesis. Size and recurrence do **not** predict lineage
association. A purely structural property -- **mutual exclusivity** -- does.

```{r libs, message=FALSE}
library(SVMC)
library(dplyr)
library(ggplot2)
```

```{r load-data}
ex <- readRDS(system.file("extdata", "salmonella_example.rds", package = "SVMC"))
M       <- ex$presence     # genomes x focal inversions (presence/absence)
serovar <- ex$serovar      # qid, serotype (independent SeqSero2 calls)
mds     <- ex$mds          # qid, MDS1, MDS2, state
ord_pct <- ex$ord_pct
ex$focal_meta
```

## Step 1 — A catalog of recurrent inversions

SVMC calls structural variants per genome and groups recurrent ones into
motifs. Here we work with five focal inversions from the *Salmonella* cohort,
ranging from 0.55 to 2.26 Mbp. Each is carried by tens to hundreds of genomes.

```{r catalog}
tibble(
  motif      = ex$focal_meta$label,
  size_mbp   = ex$focal_meta$size_mbp,
  n_carriers = colSums(M > 0)[ex$focal_meta$motif_uid]
) %>% arrange(desc(n_carriers))
```

Nothing here yet says which of these define lineages. Size and recurrence vary,
but the largest inversion (2.26 Mbp) is not the most common, and recurrence
alone does not tell us whether an inversion tracks a clade.

## Step 2 — The structural criterion: mutual exclusivity

SVMC partitions inversions by **co-occurrence**, computed from the presence
matrix alone -- no lineage or serovar information is used. Two inversions are
*mutually exclusive* when essentially no genome carries both; an inversion is
*co-occurring* when it shares carriers with another.

```{r exclusivity}
excl <- svmc_inversion_exclusivity(M)
excl
```

Two inversions fall out as **exclusive** (they share almost no carriers with any
other), and the rest as **co-occurring**. Critically, the three inversions we
will show to be lineage-defining are pairwise mutually exclusive:

```{r are-exclusive}
primary <- ex$focal_meta$motif_uid[ex$focal_meta$expected == "exclusive"]
chk <- svmc_are_mutually_exclusive(M, primary)
chk$mutually_exclusive     # TRUE: no genome carries more than one
chk$n_multiple             # 0
```

This partition is unsupervised and structural. Whatever it predicts about
biology is therefore a genuine finding, not a restatement of how the events
were chosen.

## Step 3 — Do the exclusive states predict serovar?

Now we bring in **independent** data: SeqSero2 serovar calls. For each exclusive
state we compute precision (do carriers belong to one serovar?) and recall (does
the state capture that serovar?).

```{r validation}
state_map <- c(
  "MOTIF|Structural_rearrangement|1"  = "A",   # 0.86 Mbp
  "MOTIF|Structural_rearrangement|16" = "B",   # 0.55 Mbp
  "MOTIF|Structural_rearrangement|19" = "C"    # 1.53 Mbp
)

lineage_map <- list(
  "Typhi"        = list(state = "B",
                        serovars = "Typhi"),
  "D1 serogroup" = list(state = "A",
                        serovars = c("Enteritidis","Dublin",
                                     "Gallinarum or Enteritidis","Berta")),
  "Infantis"     = list(state = "C",
                        serovars = "Infantis")
)

svmc_state_serovar_validation(M, serovar, state_map, lineage_map)
```

Each exclusive state maps to a distinct serovar with high precision. The
variation in *recall* is itself informative: the Infantis inversion is perfectly
precise but low-recall -- it marks a distinct **sublineage** of Infantis rather
than the whole serovar.

The co-occurring inversions are not shown here because they do not define a
state of their own: their carriers are largely inherited from co-occurrence with
an exclusive event, and once that is accounted for their residual signal
scatters across unrelated serovars.

## Step 4 — The same partition, visible in genomic space

Finally, the exclusive states occupy distinct territories in an ordination of
genome-wide similarity -- a third, independent view of the same partition.

```{r mds-figure, fig.width=7, fig.height=5.5}
state_pal <- c("0.86 Mbp Inversion" = "#2A9D8F",
               "0.55 Mbp Inversion" = "#E63946",
               "1.53 Mbp Inversion" = "#8338EC")

fig <- mds %>% filter(state %in% names(state_pal))

ggplot() +
  geom_point(data = mds %>% filter(!state %in% names(state_pal)),
             aes(MDS1, MDS2), color = "grey88", size = 0.7) +
  geom_point(data = fig, aes(MDS1, MDS2, fill = state),
             shape = 21, color = "white", stroke = 0.4, size = 2.4) +
  scale_fill_manual(values = state_pal, name = "Exclusive state") +
  labs(x = sprintf("MDS1 (%.1f%%)", ord_pct["MDS1"]),
       y = sprintf("MDS2 (%.1f%%)", ord_pct["MDS2"])) +
  theme_classic(base_size = 12)
```

The three mutually exclusive inversions form separate clusters; the background
(grey) and the co-occurring passengers do not carve out territories of their
own.

## What SVMC found

Putting the four steps together:

1. **Catalog** — the cohort carries several recurrent large inversions.
2. **Structural partition** — some are mutually exclusive, some co-occur
   (computed blind to lineage).
3. **Independent validation** — the exclusive inversions each predict a distinct
   serovar (72--100% precision); the co-occurring ones do not.
4. **Genomic confirmation** — the exclusive states separate cleanly in an
   ordination of genome similarity.

**Mutual exclusivity, not size or recurrence, distinguishes lineage-defining
structural variants.** The mechanistic question this raises -- why some
inversions are exclusive (plausibly, because they are alternative orientations
of the same contested genomic region) -- is a direction for future work, and one
SVMC's breakpoint annotations are positioned to address.
